\documentclass{pgnotes}

\title{XPath}

\begin{document}

\maketitle

\section{XPath}

XPath expressions allow us to easily retrieve the contents of elements and attributes in an XML document, by writing expressions.  

The XPath expression language is standardised and works similarly in many languages - Java, C\#, C++/libxml/JavaScript etc.

XPath uses \textbf{expressions} and standard \textbf{functions} to return \textbf{nodes} that are \textbf{related} to other nodes. 


\subsection{Expressions}

XPath uses path expressions to select nodes or node-sets in an XML document.
These expressions are very similar in theory to file paths or URLs:
Examples:
\begin{itemize}
\item \texttt{/Users/peadar/Intro.pdf}
\item \texttt{/Users/peadar/Documents/Intro.pdf}
\item \url{http://xyz.com/dir/file2.xml}
\item \url{http://xyz.com/dir/sd/file2.xml}
\end{itemize}


\subsection{Standard functions}

XPath includes over 100 built-in functions.
There are functions for string values, numeric values, date and time comparison, Boolean values, and more complex manipulations.
  
Functions usually \textbf{operate} on nodes or node-sets returned using an expression. 


\subsection{Nodes}

In XPath, there are seven kinds of nodes:

\begin{enumerate}
\item \textbf{Element}
\item \textbf{Attribute}
\item \textbf{Text}
\item Namespace (see later)
\item Processing instruction 
\item Comment
\item \textbf{Document}
\end{enumerate}


\subsection{Relationships}

Effective use of XPath requires a firm understanding of the relationships present between various nodes, \autoref{tab:xpath:relationships}.

\begin{table}[htbp]
  \centering
  \begin{tabularx}{1.0\linewidth}{l X}
    \toprule
    \textbf{Relationship} & \textbf{Meaning} \\
    \midrule
    Parent & Each element and attribute has one parent. \\
    Children & Element nodes may have zero, one or more children. \\
    Siblings & Nodes that have the same parent. \\
    Ancestors & A node's parent, parent's parent, etc. \\
    Descendants & A node's children, children's children, etc. \\
    \bottomrule
  \end{tabularx}
  \caption{Important relationships between nodes}
  \label{tab:xpath:relationships} 
\end{table}


\section{Basic selection expressions}

XPath uses path expressions to select nodes or node-sets in an XML document.
The node is selected by following a path or steps.
The most useful path expressions are listed in \autoref{tab:xpath-expressions}.

\begin{table}[htbp]
  \centering
  \begin{tabular}{l l}
    \toprule
    \textbf{Expression} & \textbf{Selects} \\
    \midrule
    \textit{elementname} &  all ``elementname'' elements\\
    / & from root node / previous node \\
    // & from anywhere in the document \\
    .. & parent of the current node \\
    @attribname & attribute ``attribname'' \\
    text() & text content of current node\\
    \bottomrule
  \end{tabular}
  \caption{XPath expressions}
  \label{tab:xpath-expressions}
\end{table}

\subsection{Writing path expressions}

A location path can be absolute or relative.
An absolute location path starts with a slash ( / ) and a relative location path does not.
In both cases the location path consists of one or more steps, each separated by a slash:

\begin{verbatim}
An absolute location path:

/step/step/...

A relative location path:

step/step/...
\end{verbatim}

Each step is evaluated against the nodes in the current node-set.


\subsection{Example}

Using the bookstore document, a number of XPath expressions are shown in \autoref{tab:xpath:expressions-example}. 

\begin{table}[htbp]
  \centering
  \begin{tabular}{l l}
    \toprule
    \textbf{Expression} & \textbf{Selects} \\
    \midrule
    \texttt{bookstore} & all elements named ``bookstore'' \\
    \texttt{/bookstore} & root element bookstore \\
    \texttt{bookstore/book} & child books of bookstore \\
    \texttt{//book} & all books in document \\
    \texttt{bookstore//book} & descendant books of bookstore\\
    \bottomrule
  \end{tabular}
  \caption{Expressions on \texttt{bookstore.xml} document}
  \label{tab:xpath:expressions-example}
\end{table}


\subsubsection{Selecting text content}

Note that we must use the \texttt{text()} function to get just the text content of an element, rather than the element itself:
\begin{itemize}
\item \texttt{/bookstore/book/title} = title element of each book
\item \texttt{/bookstore/book/title/text()} = title text content of each book
\end{itemize}

\section{Tooling}

\subsection{Command-line}

The \texttt{xmllint} tool can evaluate XPath expressions at the command-line.

\begin{minted}{bash}
xmllint --xpath '/route/station[amenity]' stations.xml
\end{minted}


\subsection{Python}

Python's \texttt{lxml} library can work with XPath expressions:
\begin{minted}{python}
import lxml.etree as ET

root = ET.parse('stations.xml')
for name in root.xpath('/route/station/name/text()'):
    print(name)
\end{minted}



\section{Predicates}

Predicates are used to find a \textbf{specific node} or a \textbf{node containing a specific value}.

They are embedded in square brackets after the element name to which they refer.
Examples:
\begin{itemize}
\item \texttt{/element/element[predicate]}
\item \texttt{/element[predicate]/element}
\item \texttt{/element[p1]/element[p2]}
\end{itemize}


\subsection{Posessive predicates}

Posessive predicates select elements that have matching children, \autoref{tab:xpath:posessive-predicates}. 

\begin{table}[htbp]
  \centering
  \begin{tabularx}{1.0\linewidth}{l X}
    \toprule
    \textbf{Expression} & \textbf{Selects} \\
    \midrule
    \textbf{//parent[@attribute]} &  all ``parent'' elements with  ``attribute'' \\
    \textbf{//gp/parent[child]} & all ``parent'' elements with ``child'' \\
    \textbf{//gp/parent[child]/child2} &  all ``child2'' of above... \\
    \bottomrule
  \end{tabularx}
  \caption{Posessive predicates}
  \label{tab:xpath:posessive-predicates}
\end{table}

\subsection{Value-based predicates}

Value-based predicates select elements based on whether they or their child elements or attributes match specific values, \autoref{tab:xpath:value-based-predicates}. 

  \begin{table}[htbp]
    \centering
    \begin{tabularx}{1.0\linewidth}{l X}
      \toprule
      \textbf{Expression} & \textbf{Selects} \\
      \midrule
      \texttt{//parent[@attribute='value']} &  all ``parent'' elements with ``attribute''==value \\
      \texttt{//gp/parent[child$<v$]} &  all ``parent'' elements where ``child'' element has value $<v$ \\
      \texttt{//gp/parent[child$<v$]/child2} &  all ``child2'' of above... \\
      \bottomrule
    \end{tabularx}
    \caption{Value-based predicates}
    \label{tab:xpath:value-based-predicates}
  \end{table}

\subsection{Positional predicates}

Positional predicates allow selections based on element ordering usually using XPath functions, \autoref{tab:xpath:positional-predicates}. 

\begin{table}[htbp]
  \centering
  \begin{tabularx}{1.0\linewidth}{l X}
    \toprule
    \textbf{Expression} & \textbf{Selects} \\
    \midrule
    /parent/child[$n$] & the $n$th child element of the parent. \\
    /parent/child[last()] & the last child element of the parent. \\
    /parent/child[last()-$n$] & the last but $n$ child element of the parent. \\
    /parent/child[position()$<n$] & the first $n-1$ child elements of the parent. \\
    \bottomrule
  \end{tabularx}
  \caption{Positional predicates}
  \label{tab:xpath:positional-predicates}
\end{table}

\subsection{Other predicates}

There are a lot more things you can do with predicates --- see online tutorials for more details \cite{w3-schools:2013:xpath}. 

\subsection{Combining two node sets}

To combine two node sets, we use the $\vert$ (vertical bar) operator between two separate XPath expressions.
Example: 

\begin{verbatim}
    \texttt{/path/path[predicate] $\vert$ /path2/path2[predicate]}
\end{verbatim}



\subsection{Operators}

When writing a predicate we can use any of the usual binary comparison operators, \autoref{tab:xpath:operators}.

\begin{table}[htbp]
  \centering
  \begin{tabular}{l l}
    \toprule
    \textbf{Operator} & \textbf{Meaning} \\
    \midrule
    = & equal \\
    != & not equal \\
    $<, >$ & less than, greater than \\
    $<=, >=$ & less/greater than or equal to \\
    or, and & or/and \\
    \bottomrule
  \end{tabular}
  \caption{XPath operators}
  \label{tab:xpath:operators}
\end{table}

You can use math operators in your comparisons, but \textbf{be careful of division}, \autoref{tab:xpath:math-operators}. 

\begin{table}[htbp]
  \centering
  \begin{tabular}{l l}
    \toprule
    \textbf{Operator} & \textbf{Meaning} \\
    \midrule
    $+$ & addition \\
    $-$ & subtraction \\
    * & multiplication \\
    div & division \\
    mod & modulus / division remainder\\
    \bottomrule
  \end{tabular}
  \caption{XPath operators}
  \label{tab:xpath:math-operators}
\end{table}


\section{Wildcards}

XPath wildcards can be used to select unknown XML elements, \autoref{tab:xpath-wildcards}

\begin{table}[htbp]
  \centering
  \begin{tabular}{l l}
    \toprule
    \textbf{Expression} & \textbf{Matches} \\
    \midrule
    \texttt{*} &  any element node \\
    \texttt{@*} &  any attribute node \\
    \texttt{node()} & any node of any kind \\
    \bottomrule
  \end{tabular}
  \caption{XPath wildcards}
  \label{tab:xpath-wildcards}
\end{table}





\section{Axes}

An axis defines a node-set relative to the current node.

\subsection{Direct axes}

Direct axes specify the node itself, its parent or children, or its attributes, \autoref{tab:xpath:direct-axes}. 

\begin{table}[htbp]
  \begin{tabularx}{1.0\linewidth}{l X}
    \toprule
    \textbf{Axis} & \textbf{Selects} \\
    \midrule
    self & the current node \\
    parent & the parent of the current node \\
    child & all children of the current node \\
    attribute & all attributes of the current node \\
    \bottomrule
  \end{tabularx}
  \caption{Direct axes}
  \label{tab:xpath:direct-axes}
\end{table}


\subsection{Cross-generational axes} 

Cross-generational axes allow us to specify nodes ancestral or descending nodes, \autoref{tab:xpath:cross-generational-axes}. 

\begin{table}[htbp]
  \begin{tabularx}{1.0\linewidth}{l X}
    \toprule
    \textbf{Axis} & \textbf{Selects} \\
    \midrule
    ancestor & all ancestors (parent, grandparent, etc.) of the current node \\
    ancestor-or-self & current node and all of its ancestors \\
    descendant & all descendants (children, grandchildren, etc.) of the current node \\
    descendant-or-self & current node and all descendents \\
    \bottomrule
  \end{tabularx}
  \caption{Cross-generational axes}
  \label{tab:xpath:cross-generational-axes}
\end{table}




\subsection{Positional axes} 

Positional axes allow us to select nodes before or after other nodes in the document, \autoref{tab:xpath:positional-axes}. 

\begin{table}[htbp]
  \begin{tabularx}{1.0\linewidth}{l X}
    \toprule
    \textbf{Axis} & \textbf{Selects} \\
    \midrule
    following & everything in the document after the closing tag of the current node \\
    following-sibling & all siblings after the current node \\
    preceding &  all nodes  before  current node, except ancestors, attribute nodes and namespace nodes \\
    preceeding-sibling & all siblings before the current node \\
    \bottomrule
  \end{tabularx}
  \caption{Positional axes}
  \label{tab:xpath:positional-axes}
\end{table}

\subsection{Writing location path expressions}

With predicates and axes, a full step consists of:

\begin{itemize}
\item an axis (defines the tree-relationship between the selected nodes and the current node)
\item a node-test (identifies a node within an axis)
\item zero or more predicates (to further refine the selected node-set)
\end{itemize}

The syntax for a location step is:
\begin{verbatim}
axisname::nodetest[predicate]
\end{verbatim}


\subsection{Example}

Here are some XPath expressions using axes for the bookstore.xml file:

\begin{table}[htbp]
  \begin{tabularx}{1.0\linewidth}{l X}
    \toprule
    \textbf{Expression} & \textbf{Selects} \\
    \midrule
    child::book & 	 all book nodes that are children of the current node \\
    attribute::lang &	 the lang attribute of the current node \\
    child::* &	 all element children of the current node \\
    attribute::* &	 all attributes of the current node \\
    child::text() &	 all text node children of the current node \\ 
    child::node() &	 all children of the current node \\ 
    descendant::book &	 all book descendants of the current node \\ 
    ancestor::book &	 all book ancestors of the current node \\ 
    ancestor-or-self::book &	 all book ancestors of the current node - and the current as well if it is a book node \\
    child::*/child::price &	 all price grandchildren of the current node\\
    \bottomrule
  \end{tabularx}
  \caption{XPath axes expressions example}
  \label{tab:xpath-axes-example}
\end{table}


\section{XQuery}

XQuery is to XML what SQL is to database tables.
XQuery was designed to \textbf{query} rather than \textit{navigate} XML data.

\subsection{Syntax}

XQuery is built on top of \textbf{XPath}, and uses the \textbf{same expressions and predicates},  
augmented by the queries made up of by the so-called \textbf{FLWOR} syntax, \autoref{tab:flwor}.

\begin{table}[htbp]
  \begin{tabularx}{1.0\linewidth}{l X}
    \toprule
    \textbf{Component} & \textbf{Function} \\
    \midrule
    \textbf{f}or & creates a sequence of nodes \\
    \textbf{l}et & binds a sequence to a variable \\
    \textbf{w}here & filters nodes on a boolean expression \\
    \textbf{o}rder by & sorts the nodes \\
    \textbf{r}eturn & evaluated once per node \\
    \bottomrule
  \end{tabularx}
  \caption{FLWOR syntax}
  \label{tab:flwor}
\end{table}


\subsection{Rules}

\begin{enumerate}
\item XQuery is case-sensitive
\item XQuery elements, attributes, and variables must be valid XML names
\item An XQuery string value can be in single or double quotes
\item Local variables (usually to represent the current node) are preceeded by the \$ character. 
\item XQuery comments are delimited by (: and :), e.g. (: XQuery Comment :)
\end{enumerate}


\subsection{Examples}

We will use the following XML document:

\inputminted{xml}{xpath_code/books.xml}


\subsubsection{Basic selection}

We are asked to:
\begin{quotation}
Select all the title elements under the book elements that are under the bookstore element that have a price element with a value that is higher than 30.
\end{quotation}

Write down a suitable XPath expression for this: 

\vspace{\fill}

This same query could be expressed using the FLWOR syntax as follows:

\begin{minted}{xquery}
for $x in /bookstore/book
where $x/price > 30.00
return $x/title
\end{minted}

Obviously, the XPath expression is more concise. 
However, we have a problem if we want the returned element list to be ordered or otherwise modified:

\begin{minted}{xquery}
for $x in /bookstore/book
where $x/price > 30.00
order by $x/title
return $x/title
\end{minted}

We can simplify by using XPath predicates instead of (or in addition to) the WHERE clause: 

\begin{minted}{xquery}
for $x in /bookstore/book[price>30]
order by $x/title
return $x/title
\end{minted}

\subsection{Counting loop iterations}

The at keyword can be used to count the iteration:

\begin{minted}{xquery}
for $x at $i in /bookstore/book/title
return {$i}. {data($x)}
\end{minted}

\subsection{Multiple order by clauses}

The order by clause is used to specify the sort order of the result. Here we want to order the result by category and title:

\begin{minted}{xquery}
for $x in /bookstore/book
order by $x/@category, $x/title
return $x/title
\end{minted}

\bibliographystyle{plainnat}
\bibliography{bibliography}

\end{document}



